---
title: "Dynamic Mode Decomposition as a Linear Model for Spatio-Temporal Systems"
author: "Jeremy Roberts"
date: "December 1, 2019"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## The Problem

The aim is to construct linear statistical models for some process of the form

$$
  \frac{d\mathbf{z}}{dt} = \mathbf{f}(\mathbf{z}) \, ,
$$
observed at fixed locations in time to produce the "snapshot" matrices

$$
  \mathbf{Z}_- =  [ \mathbf{z}_{1},\mathbf{z}_{2},\ldots, \mathbf{z}_{n-1}] \quad \text{and} \quad 
  \mathbf{Z}_+ =  [\mathbf{z}_{2},\mathbf{z}_{3}, \ldots,\mathbf{z}_{n}] \, ,
$$

where $\mathbf{z}_i = \mathbf{z}(t_i)$ and $t_i - t_{i-1} = \Delta\, , \forall i$.

## Dynamic Mode Decomposition

Koopman says there exists some infinite-dimensional operator $\mathcal{A}$ such that $\mathcal{A}\mathbf{z} = \mathbf{f}(\mathbf{z}) = \frac{d\mathbf{z}}{dt}$.
We're practical and seek the finite-dimensional operator (matrix) $\mathbf{A}$ such that

$$
  \mathbf{Z}_+ \approx \mathbf{A}\mathbf{Z}_- \, .
$$
With eigenpairs $\boldsymbol{\phi}_i$ and $\lambda_i$ of $\mathbf{A}$, 

$$
\boxed{\mathbf{z}(t) \approx \sum^{}_{i} b_i \boldsymbol{\phi}_i e^{\frac{\log(\lambda_i) t}{\Delta}} } \, .
$$

## Low-Rank Approximation

The best-fit operator $\mathbf{A} = \mathbf{Z}_+ \mathbf{Z}_-^{\dagger}$ is too unwieldy, so use low-rank, singular-value decomposition

$$ 
  \mathbf{Z}^{\dagger}_- \approx \mathbf{V}_r\boldsymbol{\Sigma}^{-1}_r \mathbf{U}^*_r 
$$
to define the low-rank operator

$$
  \tilde{\mathbf{A}} = \mathbf{U}^*_r \mathbf{A} \mathbf{U}_r = \mathbf{U}^*_r \mathbf{Z}_+ \, .
$$
The eigendecomposition $\tilde{\mathbf{A}}\mathbf{W} = \mathbf{W}\boldsymbol{\Lambda}$ yields eigenvalues $\lambda_i$ and eigenvectors $\boldsymbol{\Phi} =  \mathbf{Z}_+ \mathbf{V}_r\boldsymbol{\Sigma}_r^{-1}\mathbf{W}$ of $\mathbf{A}$.

Because modes are complex, (1) keep Re and Im parts of pairs separately or (2) keep only Re component.

## Demonstration Problem

To illustrate, maximum temperatures at 132 weather stations in 07/93 were modeled using NOAA data packaged for [Spatio-Temporal Statistics with R
](https://spacetimewithr.org/).


```{r tidy=TRUE,message=FALSE,warning=FALSE}
library("dplyr")
library("STRbook")
data("NOAA_df_1990", package = "STRbook")
df <- filter(NOAA_df_1990, year==1993, month==7, proc=="Tmax")
keep <- c()
for (station in unique(df$id)){
  if (length(filter(df,id==station)$julian)==31){
    keep <- append(keep, station)
  }
}
df <- filter(df, id%in%keep)
df <- mutate(df, t=df$julian-min(df$julian))
df <- subset(df, select=c(t, id, z, lat, lon))
num_locs <- length(unique(df$id))
num_time <- length(df$id)/num_locs
Z <- t(matrix(as.numeric(df$z), num_time, num_locs, byrow = FALSE))
library(ggplot2)
heat <- expand.grid(time=1:num_time, location=1:num_locs)
heat$z <- df$z
ggplot(heat, aes(location, time, fill=z)) + geom_tile() + scale_fill_continuous(limits=c(60, 110), type = "viridis", option="inferno")
```

## Inspecting DMD modes

```{r tidy=TRUE,message=FALSE,warning=FALSE}
source("dmd.R")
library(latex2exp)
library(ggplot2)
require(gridExtra)

dmd <- compute_dmd(Z, rank=0, keep_pairs=0)
tmp <- sort(filter(df, t==1)$lat, index.return=TRUE)
ix <- tmp$ix
lats <- tmp$x
modes <- Re(dmd$Phi[ix, ])
modes <- modes %*% diag(1/sqrt(colSums(modes^2)))
tmp_df = data.frame(lats=lats, Phi=modes)
```

Here,  four DMD modes using optimal rank (i.e., `r dmd$rank`) and discarding imaginary components of complex-conjugate pairs:

```{r tidy=TRUE,message=FALSE,warning=FALSE, fig.width=6,fig.height=3.5}
plots = list()
for (i in 1:4)
{
  plots[[i]] <-  ggplot(tmp_df, aes_string("lats",  sprintf("Phi.%i", i))) + 
                 geom_line(alpha=0.2) +
                 geom_point() +
                 xlab("latitude") + 
                 ylab(TeX(sprintf("$\\phi_{%i}$", i))) +
                 labs(title=TeX(sprintf("$\\omega_{%i} = %.4f + %.4f \\, i$", i, Re(dmd$omega[i]), Im(dmd$omega[i]))))
}
q <- grid.arrange(grobs=plots, ncol=2)

```

Note the near-zero frequency of mode 1, which indicates a nearly time-invariant contribution.

## Reconstructing $\mathbf{Z}$ with DMD Basis


```{r tidy=TRUE,message=FALSE,warning=FALSE}
require(gridExtra)

models = list()
# Make DMD models and predictions
for (r in c(5, 7, -1)){
  tmp <- lm_with_dmd(df, rank=r, keep_pairs = 2)
  tmp$pred <- predict_with_dmd(df, tmp$model, tmp$dmd)
  tmp$desc <- sprintf("DMD, rank=%i", tmp$dmd$rank)
  models[[length(models)+1]] <- tmp
}
for (i in 1:3)
{
  desc <- "z ~ (t+lat+lon)"
  if (i > 1){desc <- sprintf("z ~ (t+lat+lon)^%i",i)}
  tmp  <- list(model=lm(formula(desc), data=df))
  tmp$pred <- predict(tmp$model, interval="prediction")
  tmp$desc <- desc
  models[[length(models)+1]] <- tmp
}

# Given data, the model, and the prediction, make a plot of the prediction,
# make a plot of the prediction error, and compute a few key quantities.
p = list()
for (i in 1:length(models)){
  m <- models[[i]]

  heat <- expand.grid(time=1:num_time, location=1:num_locs)
  heat$z <-m$pred[, 1]
  p[[length(p) + 1]] <- ggplot(heat, aes(location, time, fill=z)) +
                        geom_tile() + theme_void() + labs(title=m$desc) +
                        scale_fill_continuous(limits=c(60, 110), type = "viridis", option="inferno")
  
  aic <- AIC(m$model)
  desc <- sprintf("AIC = %.2f", aic)
  heat$error <- m$pred[, 1] - df$z
  
  p[[length(p) + 1]] <- ggplot(heat, aes(location, time, fill=error)) + 
                        geom_tile() + scale_fill_distiller(palette = "Spectral", limits=c(-25, 25)) + theme_void() +
                        labs(title=desc) 
  
  p[[length(p) + 1]]  <- ggplot(filter(cbind(df, m$pred), id==14943), aes(t, z)) +
       geom_point() +
       geom_line(aes(y = fit), color="blue") +
       geom_line(aes(y = upr), color = "red", linetype = "dashed") +
       geom_line(aes(y = lwr), color = "red", linetype = "dashed") 
}

q <- grid.arrange(grobs=p[1:9], ncol=3)
```

## Reconstructing $\mathbf{Z}$ with Poly. Basis

```{r}
q <- grid.arrange(grobs=p[10:18], ncol=3)
```


## Predicting $\mathbf{Z}$ with DMD Basis

Now, use only first 28 observations.

```{r tidy=TRUE,message=FALSE,warning=FALSE}
require(gridExtra)

df_partial <- filter(df, t < 28)

models2 = list()
# Make DMD models and predictions
for (r in c(5, 7, -1)){
  tmp <- lm_with_dmd(df_partial, rank=r,keep_pairs = 2)
  tmp$pred <- predict_with_dmd(df, tmp$model, tmp$dmd)
  tmp$desc <- sprintf("DMD, rank=%i", tmp$dmd$rank)
  models2[[length(models2)+1]] <- tmp
}
for (i in 1:3)
{
  desc <- "z ~ (t+lat+lon)"
  if (i > 1){desc <- sprintf("z ~ (t+lat+lon)^%i",i)}
  tmp  <- list(model=lm(formula(desc), data=df_partial))
  tmp$pred <- predict(tmp$model, df, interval="prediction")
  tmp$desc <- desc
  models2[[length(models2)+1]] <- tmp
}

# Given data, the model, and the prediction, make a plot of the prediction,
# make a plot of the prediction error, and compute a few key quantities.
p = list()
for (i in 1:length(models2)){
  m <- models2[[i]]

  heat <- expand.grid(time=1:num_time, location=1:num_locs)
  heat$z <-m$pred[, 1]
  p[[length(p) + 1]] <- ggplot(heat, aes(location, time, fill=z)) +
                        geom_tile() + theme_void() + labs(title=m$desc) +
                        scale_fill_continuous(limits=c(60, 110), type = "viridis", option="inferno")
  
  aic <- AIC(m$model)
  desc <- sprintf("AIC = %.2f", aic)
  heat$error <- m$pred[, 1] - df$z
  p[[length(p) + 1]] <- ggplot(heat, aes(location, time, fill=error)) + 
                        geom_tile() + scale_fill_distiller(palette = "Spectral", limits=c(-25, 25)) + theme_void() +
                        labs(title=desc) 

  p[[length(p) + 1]]  <- ggplot(filter(cbind(df, m$pred), id==14943), aes(t, z)) +
       geom_point() +
       geom_line(aes(y = fit), color="blue") +
       geom_line(aes(y = upr), color = "red", linetype = "dashed") +
       geom_line(aes(y = lwr), color = "red", linetype = "dashed") 
}

q <- grid.arrange(grobs=p[1:9], ncol=3)#, theme(plot.margin=margin(250,10,10,10)))
```

## Predicting $\mathbf{Z}$ with Poly. Basis

```{r}
q <- grid.arrange(grobs=p[10:18], ncol=3)
```

## A Bit on Model Selection

```{r}
library(leaps)

aic <- rep(0, 30)
aic2 <- rep(0, 30)
ranks <- rep(0, 31)
for (r in 1:30){
  m <- lm_with_dmd(df, rank=r, keep_pairs=0)
  aic[r] <- AIC(m$model)
  ranks[r] <- r-sum(is.na(m$model$coefficients))
  m <- lm_with_dmd(df, rank=r, keep_pairs=2)
  aic2[r] <- AIC(m$model)
}

m <- lm_with_dmd(df, rank=-1, keep_pairs=2)
tmp_df <- df_add_dmd(subset(df, select=c(z, t)), m$dmd) 
b <- regsubsets(z ~. - t, intercept=FALSE, data=tmp_df, nvmax=30)
rs <- summary(b)
num_models <- dim(rs$which)[1]
aic3 <- rep(0.0, num_models)
for (i in 1:num_models){
  aic3[i] <- AIC(lm(z~.-1, data=subset(tmp_df, select=c(z=TRUE, t=FALSE, rs$which[i,]))))
}

plot(1:30, aic, type="b", pch=19, col="black", xlab="DMD rank", ylab="AIC", ylim=c(min(aic,aic2), max(aic, aic2)))
lines(1:30, aic2, pch=18, col="red", type="b", lty=2)
lines(1:30, aic3, pch=17, col="blue", type="b", lty=2)
legend(1, 30000, legend=c("eliminate", "keep", "30"),  col=c("black", "red", "blue"), lty=1:2, cex=0.8, pch=c(19,18,17))


```
