---
title: "Dynamic Mode Decomposition as a Linear Model for Spatio-Temporal Systems"
author: "Jeremy Roberts"
date: "December 1, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This notebook includes the basic analysis shown in the final report.  The report includes the necessary theory (equations, etc.), while this notebook puts the code into action.

# Load Data

First, load the maximum temperature data for July of 1993:

```{r tidy=TRUE,message=FALSE,warning=FALSE}
library("dplyr")
library("STRbook")
data("NOAA_df_1990", package = "STRbook")
df <- filter(NOAA_df_1990, year==1993, month==7, proc=="Tmax")
```

Now, we need to clean up the data to include only those stations that have measurements for each day of the month.  This greatly simplifies the DMD analysis, which is based on a dense [space, time] snapshot matrix.

There are 31 days in July, so search for all those station id's that have 31 measurements and filter the data to include only those.

```{r tidy=TRUE,message=FALSE,warning=FALSE}
keep <- c()
for (station in unique(df$id)){
  if (length(filter(df,id==station)$julian)==31){
    keep <- append(keep, station)
  }
}
df <- filter(df, id%in%keep)
```

The time coordinate is a cardinal month index.  Here, add a new column for time that translates the time to start at zero:

```{r}
df <- mutate(df, t=df$julian-min(df$julian))
```

Finally, to clean things up a bit, keep only `t`,  `id` (the station number), `z` (the maximum temperature), `lat` (latitude), and `lon` (longitude):
```{r}
df <- subset(df, select=c(t, id, z, lat, lon))
head(df)
```

# Construct Snapshot Matrix

DMD is applied to a matrix $\mathbf{Z} \in \mathbb{R}^{m,n}$, or

```{r tidy=TRUE,message=FALSE,warning=FALSE}
num_locs <- length(unique(df$id))
num_time <- length(df$id)/num_locs
Z <- t(matrix(as.numeric(df$z), num_time, num_locs, byrow = FALSE))
```

One way to view this data is by treating the elements of $\mathbf{Z}$ as pixels of an image (or a "heatmap").  Here, the temperature is shown on a 2-D grid of location (an arbirary index for each weather station) and day (in July):

```{r tidy=TRUE,message=FALSE,warning=FALSE}
library(ggplot2)
heat <- expand.grid(time=1:num_time, location=1:num_locs)
heat$z <- df$z
ggplot(heat, aes(location, time, fill=z)) + geom_tile() + scale_fill_viridis_c(option="inferno")
```

# A DMD-generated Basis for Statistical Linear Modeling 

Given $\mathbf{Z}$, $\mathbf{Z}_- = \mathbf{Z}_{:,1:n-1}$, and $\mathbf{Z}_+ = \mathbf{Z}_{:, 2:n}$, the DMD algorithm employs the SVD to compute a rank-$r$ operator $\mathbf{A}_r$ that approximately satisfies $\mathbf{Z}_+ = \mathbf{A}_r\mathbf{Z}_-$.  The eigenvectors $\boldsymbol{\phi}_i$ of $\mathbf{A}_r$ are the DMD "modes", while  the corresponding eigenvalues $\lambda_i$ yield the DMD frequencies $\omega_i = \log \lambda_i / \Delta_t$ for observations separated by a fixed $\Delta_t$ in time.

To support modeling of $\mathbf{z}(t)$ given the observed data, construction and use of a linear model based on `lm` requires a few steps:

  1. Generate the DMD modes $\boldsymbol{\phi}_i$ and frequences $\omega_i$ given known observations and times.
  2. Construct spatio-temporal DMD basis vectors $\boldsymbol{\psi}_i$ using the same observations and times.
  3. Use `lm` to fit the model $\mathbf{z} \sim \mathcal{N}(\boldsymbol{\Psi}\boldsymbol{\beta}, \sigma^2\mathbf{I})$, where the DMD basis vectors are the covariates.  This requires that a dataframe be generated with columns for each of those basis vectors. 
  4. Predictions should be made with `lm` for possibly different times than were used in the original fitting.  In other words, the basis vectors may need to be reconstructed at new times with which a new dataframe is constructed.
  
To support these steps, several functions are implemented in `dmd.R`:

  - `compute_svd(Z, r)` computes the SVD of a matrix $\mathbf{Z}$ with a specified rank $r$.  If $r > 0$, the leading $r$ singular vectors and values are retained.  If $r = 0$, the optimal rank is selected (based on the criterion of Davish and Donoho).  If $r < 0$, then the maximum possible rank is used (i.e., $\text{max}(n, m)$ for $\mathbf{Z}\in \mathbb{R}^{m\times n}$).  A `list` with entries `u`, `v`, and `d` is returned, just like the R-function `svd`.
  - `compute_dmd(Z, rank, delta.t)` computes the DMD modes and frequencies given $\mathbf{Z}$, a desired rank, and a fixed $\Delta_t$.  The rank is the same used for the SVD.  However, because complex-conjugate pairs are reduced to their real components, the final rank of the DMD modes may be reduced.  A list with entries `Phi`, `omega`, `rank`, and `delta.t` is returned, where this `rank` is the final one.
  - `compute_dmd_basis(dmd, t)` computes the DMD basis vectors given the modes `dmd$Phi`, frequencies `dmd$omega`, and times `t`.  The matrix `\boldsymbol{\Psi} \in \mathbb{R}^{mn, r}$ is returned, where $r$ is the number of DMD modes.
  - `df_add_dmd(df, dmd)` inserts a DMD basis computed using `dmd` into a dataframe `df`.  The dataframe must have observations organized in a regular space-time grid (with time as the inner variable).  Times may be arbitrary (relative to a starting point of 0), but spatial points must be consistent with those of the DMD modes.  Columns added are `Psi.1`, `Psi.2`, ..., `Psi.R`.
  - `lm_with_dmd(df, rank)` produces a linear model using the equivalent of `z ~ Psi.1 + Psi.2 + ... + Psi.R - 1`, i.e., the intercept is removed because `Psi.1` captures the "average" dynamics.  Returns a list with `model` and `dmd` attributes.
  - `predict_with_dmd(df, dmdmodel)` is given (a possibly new) dataframe and list `dmdmodel` from `lm_with_dmd`.  Returns the result of `predict` with 95% prediction intervals.
  
  
# Inspecting DMD Modes

With the helper functions in place, it is useful to examine what goes into a DMD basis.  Here, the full SVD and the optimally truncated SVD are computed to show where the "optimal" criterion sits for this data:

```{r}
source("dmd.R")
library(latex2exp)

svd_f <- compute_svd(Z, rank=-1)
svd_t <- compute_svd(Z, rank=0)
plot(svd_f$d, log="y", col="black", type="o", , xlab="", ylab="")
title(xlab=TeX('$i$'), ylab=TeX('$\\sigma_i$'))
r <- length(svd_t$d)
text(x=10, y=svd_t$d[r], labels = sprintf("optimal rank is %i", r), pos = 3., adj=c(0,0), col="red")
points(svd_t$d, col="red", type="o")
```

The optimal rank is `r r`.

With this rank, compute the DMD modes and amplitudes.  

```{r}
dmd <- compute_dmd(Z, rank=0)
lats <- filter(df, t==1)$lat #
plot(lats, dmd$Phi[, 1])
for (r in 2:dmd$rank){
  points(lats, dmd$Phi[, 2])
}
```
  

